A Theory for Distributed Database Transactions

### 2019-06-15 2018_Transactions/articleTransactions.tex
repair mistakes

\section{History}
\label{sct:History}
	A treatment of time lays the foundation of our theory of distributed information systems (DIS).
	Because of the distributed nature of a DIS,
	we may not know in which order two events\footnote{Cambridge Dictionary defines ``event'' as: anything that happens} have happened.
	For this reason we describe the history of a DIS as a partially ordered relation of events.

@@

	We represent the ``history of a DIS up to some point in time'' by a relation $\id{prec}$ (pronounce: {\em precedes}),
	being a set of pairs from $\mathbb E$.
\[\id{prec}\subseteq\mathbb E\times\mathbb E\]
	Let $e_1\ \id{prec}\ e_2$ represent the fact that event $e_1$ has happened before event $e_2$,
	making $\id{prec}$ a relation on events in $\mathbb E$.
	We require:
\begin{eqnarray}
	&&(e,e)\notin\id{prec}\label{req:prec irreflexive}\\
	&&(e_1,e_2)\in\id{prec}^+\ \wedge\ (e_2,e_1)\in\id{prec}^+\ \Rightarrow\ e_1=e_2\label{req:precPlus asy}
\end{eqnarray}
	The events in the DIS, of which $\id{prec}$ represents the history, are 
\begin{eqnarray}
	&&\id{events}(\id{prec})\ =\ \{e_1|(e_1,e_2)\in\id{prec}\}\cup\{e_2|(e_1,e_2)\in\id{prec}\}\label{req:events}
\end{eqnarray}
	Requirement~\ref{req:prec irreflexive} excludes the situation that $e$ has happened before itself.
	To represent time faithfully, equation~\ref{req:precPlus asy} requires that the transitive closure $\id{prec}^+$ is antisymmetric.
	This makes $\id{prec}^*$ a reflexive, transitive, and antisymmetric relation, which lets us use it as a temporal order between events.
	We call $\id{prec}$ ``a history'' because it represents all events that have happened up to a certain point in time and the (partial) order between them.
	Let $\mathbb H$ be the set of all conceivable histories, meaning that every $\id{prec}\in\mathbb H$.

	In a history $\id{prec}$, an event $e$ that precedes another event is called a {\em past} event,
	which we define as \ $\exists e_1:\ e\ \id{prec}\ e_1$.
	All other events are called {\em current} events in $\id{prec}$.
	Function $\id{current}$ computes the current events in history $\id{prec}$.
	Function $\id{initial}$ computes the initial events, which are the events without preceding events.
\begin{eqnarray}
	\id{current}(\id{prec})&=&\{e |\ \exists e_1: (e_1,e)\in\id{prec}\ \wedge\ \neg\exists e_2: (e,e_2)\in\id{prec}\}\\
	\id{initial}(\id{prec})&=&\{e |\ \exists e_1: (e,e_1)\in\id{prec}\ \wedge\ \neg\exists e_2: (e_2,e)\in\id{prec}\}
\end{eqnarray}

@@

	Let $\mathbb T$ be the set of all conceivable traces.
	With \id{prec} being a history, we define a trace $(e_1,e_2)_\id{prec}$ as follows:
\begin{definition}[Traces]
\label{def:Traces}
\item   $(e_1,e_2)_\id{prec}\in\mathbb T\ \Leftrightarrow\ (e_1,e_2)\in\id{prec}^*$
\end{definition}
	If it is obvious from the context which history ($\id{prec}$) is meant, we may write $(e_1,e_2)$ rather than $(e_1,e_2)_\id{prec}$.
	A trace algebra is a triple $\la\mathbb T,+,0\ra$,
	where the operator $+$ (concatenation) and empty trace $0$ satisfy the axioms of trace algebras:
\begin{eqnarray}
	x+(y+z)&=&(x+y)+z\label{axi:associative}\\
	0+x&=&x+0\label{axi:lr-identity}\\
	x+y\ =\ x+z&\Rightarrow&y=z\label{axi:left cancellation}\\
	x+z\ =\ y+z&\Rightarrow&x=y\label{axi:right cancellation}\\
	x+z\ =\ 0&\Rightarrow&x=0\label{axi:no inverse}
\end{eqnarray}
	In terms of universal algebra,
	equations~\ref{axi:associative} and~\ref{axi:lr-identity} state that a trace algebra is a {\em monoid}.
	Note that $+$ is {\em not} commutative.
	Equations~\ref{axi:left cancellation} and~\ref{axi:right cancellation} state that $+$ is {\em cancellative}.
	Equation~\ref{axi:no inverse} implies that there is no inverse.
	It makes the cancellative monoid into a trace algebra.

	For showing that definition~\ref{def:Traces} satisfies the axioms of trace algebra,
	we define the empty trace ($0$) and the trace concatenation ($+$):
\begin{definition}[Empty trace]
\label{def:Empty trace}
\item   $\forall e: 0_\id{prec}\ =\ (e,e)_\id{prec}$
\end{definition}
	We denote the empty trace as $0$ and define $(e_1,e_1)=0=(e_2,e_2)$.

	If $(e_1,e)_\id{prec}$ and $(e,e_2)_{\id{prec}'}$ are traces then
	we define the concatenation operator $+:\mathbb T\rightarrow\mathbb T\rightarrow\mathbb T$ by
\begin{definition}[Concatenation of traces]
\label{def:concatenation}
\item   $(e_1,e)_\id{prec}+(e,e_2)_{\id{prec}'}=(e_1,e_2)_{\id{prec}\cup\id{prec}'}$
\end{definition}
	Note that the definition of $+$ is partial, because $(e_1,e_2)_\id{prec}+(e_3,e_4)_{\id{prec}'}$ is not defined unless $e_2=e_3$.
	Nevertheless, the $+$ operator is closed: the result of composing two traces is always a trace.
	(proof needed)

@@

\section{State}
\label{sct:State}
	Typically, distributed information systems are stateful.
	Every event in the history of a DIS may cause a different (local) state.
	(The distributed nature of a DIS means that states are local.)
	Let $\mathbb S$ be the set of all conceivable states.
	So we need a function $\id{state}_\id{prec}:\mathbb E\rightarrow\mathbb S$ that assigns one state to every event $e$ in history $\id{prec}$.
	State $\id{state}_\id{prec}(e)$ represents the state immediately after $e$ has happened.
	If the history $\id{prec}$ is obvious from the context, we may write $\id{state}(e)$ rather than $\id{state}_\id{prec}(e)$.
	We assume an empty state, $\emptyset$, which is the state immediately before an initial event has happened.
	The global state of the system, if relevant, can be computed by $\{\id{state}(e) | e\in\id{current}(\id{prec})\}$.
	Locally we work with one specific current event, $e$, with $\id{state}_\id{prec}(e)$ representing the local current state.

@@

	We need operators to compute a new state ($s_2$) from a given state ($s_1$).
	In fact, we need two operators, $\oplus:\mathbb S\times\mathbb D\rightarrow\mathbb S$ and $\ominus:\mathbb S\times\mathbb S\rightarrow\mathbb D$, such that
\begin{equation}
	s_2\ =\ s_1\oplus(s_2\ominus s_1)
\label{req:oplus and ominus}
\end{equation}
	Let us call $s_2\ominus s_1$ the difference between states $s_2$ and $s_1$.
	The following equations represent the laws about state.
	They must be kept true all the time in a DIS.
\begin{eqnarray}
	\delta_\id{prec}(e_1,e_2)&=&\id{state}_\id{prec}(e_2)\ominus\id{state}_\id{prec}(e_1)\\
	e_1\ \id{prec}\ e_2&\Rightarrow&\id{state}_\id{prec}(e_2)\ =\ \id{state}_\id{prec}(e_1)\oplus\delta_\id{prec}(e_1,e_2)
\end{eqnarray}

@@

	A transaction is denoted as $\id{act}_{e_1,e_2,\Delta}$, which is a mapping from one DIS to another
\begin{equation}
\label{def:transaction}
\begin{array}{rl}
	\multicolumn{2}{l}{e_2\not\in\id{events}(\id{prec})\ \wedge\ (e_1,e_2)\in\id{prec}'^*\ \Rightarrow}\\
	&\id{act}_{e_1,e_2,\Delta}(\id{prec},\delta_\id{prec})\ =\ (\id{prec}',\delta'_{\id{prec}'})\\
	&\text{where}\\
	&\ \ \begin{array}[t]{rcll}
		\delta'_{\id{prec}'}(e,e')&=&\Delta&\text{if }(e,e')=(e_1,e_2)\\
		\delta'_{\id{prec}'}(e,e')&=&\delta_\id{prec}(e,e')&\text{otherwise}\\
		\end{array}
\end{array}
\end{equation}
	The state function can be computed recursively as follows:
\begin{equation}
\begin{array}{rcll}
	\id{state}_\id{prec}(e)&=&\id{state}_\id{prec}(e_1)\oplus\delta(e_1,e)&\text{if }(e_1,e)\in\id{prec}\\
	                       &=&\emptyset&\text{otherwise}
\end{array}
\end{equation}

	Now let us define the classical characteristic properties of transactions.
	Let us start with atomic.
	Atomicity expresses that a trace of multiple events can be treated as a single pair $\in\id{prec}$.
	To define this, we need an auxiliary function
	$\id{atomize}:\mathbb E\times\mathbb E\times\mathbb H\rightarrow\mathbb H$,
	which replaces a trace $(e_1,e_2)_\id{prec}$ by the pair $(e_1,e_2)$
	in a history $\id{prec}$.
\begin{eqnarray}
	\begin{array}[t]{r@{}l@{}l}
	\multicolumn{3}{l}{\id{atomize}(e_1,e_2,\id{prec})}\\
	   \ =\ \id{prec}-&(&\{ (e',e)\in\id{prec}| e_1\ \id{prec}^*\ e\ \wedge\ e\ \id{prec}^*\ e_2\}\ \cup\\
			  & &\{ (e,e')\in\id{prec}| e_1\ \id{prec}^*\ e\ \wedge\ e\ \id{prec}^*\ e_2\}\\
			  &)\\
		\cup&&\{ (e_1,e_2) \}\\
	\end{array}
\end{eqnarray}

@@

	Now let us turn to consistency.
	This property means that a transaction preserves truth of the state.
	We need a predicate $\mathcal P\{\mathbb S\}$ to express that a state is consistent.
\begin{definition}[Consistency]
\item A transaction $\id{act}_{e_1,e_2,\Delta}$ is consistent w.r.t. $\mathcal P$ iff:
\[\mathcal P\{\id{state}_\id{prec}(e_1)\}\ \Rightarrow\ \mathcal P\{\id{state}_\id{prec}(e_2)\}\]
\label{def:Consistency}
\end{definition}

	The third property is isolation.
	A transaction $\id{act}_\id{prec}(e_1,e_2)$ is isolated, i.e. performed in isolation, if it does not affect any other transaction in the system.
	This means that the only event in trace $(e_1,e_2)_\id{prec}$ that precedes events outside that trace is $e_1$.
	Likewise, the only event in trace $(e_1,e_2)_\id{prec}$ that is preceded by events outside that trace is $e_2$.
	We can now define isolation as the property that the traces of the ``atomized'' history
	are the same as the atomization of all traces in that history.
\begin{definition}[Isolation]
\item A transaction $\id{act}_{e_1,e_2,\Delta}$ works isolated in history $\id{prec}$ iff:
\label{def:Isolation}
\begin{eqnarray}
	(\id{atomize}(e_1,e_2,\id{prec}))^*&=&\id{atomize}(e_1,e_2,\id{prec}^*)
\end{eqnarray}
\end{definition}

>>>>

\section{History}
\label{sct:History}
	A treatment of time lays the foundation of our theory of distributed information systems (DIS).
	Because of the distributed nature of a DIS,
	we may not know in which order two events\footnote{Cambridge Dictionary defines ``event'' as: anything that happens} have happened.
	For this reason a partial order is used to describe the history of a DIS.

@@

	We represent the ``history of a DIS up to some point in time'' by a relation $\id{prec}$ (pronounce: {\em precedes}),
	being a set of pairs from $\mathbb E$.
\[\id{prec}\subseteq\mathbb E\times\mathbb E\]
	Let $e_1\ \id{prec}\ e_2$ represent the fact that event $e_1$ has happened before event $e_2$,
	making $\id{prec}$ a relation on events in $\mathbb E$.
	We require:
\begin{eqnarray}
	&&\id{prec}\ \subseteq\ \overline{I}\label{req:prec irreflexive}\\
	&&\id{prec}^+\cap{\flip{\id{prec}}}^+\ \subseteq\ I\label{req:precPlus asy}
\end{eqnarray}
	The events in the DIS, of which $\id{prec}$ represents the history, are 
\begin{eqnarray}
	\id{events}(\id{prec})&=&(V;\id{prec}\cap I)\ \cup\ (I\cap\id{prec};V)
\end{eqnarray}
	Requirement~\ref{req:prec irreflexive} excludes the situation that $e$ has happened before itself.
	To represent time faithfully, equation~\ref{req:precPlus asy} requires that the transitive closure $\id{prec}^+$ is antisymmetric.
	This makes $\id{prec}^*$ a reflexive, transitive, and antisymmetric relation, which lets us use it as a temporal order between events.
	We call $\id{prec}$ ``a history'' because it represents all events that have happened up to a certain point in time and the (partial) order between them.
	Let $\mathbb H$ be the set of all conceivable histories, meaning that every $\id{prec}\in\mathbb H$.

	In a history $\id{prec}$, an event $e$ that precedes another event is called a {\em past} event,
	which we define as \ $\exists e_1:\ e\ \id{prec}\ e_1$.
	All other events are called {\em current} events in $\id{prec}$.
	Function $\id{current}$ computes the current events in history $\id{prec}$.
	Function $\id{initial}$ computes the initial events, which are the events without preceding events.
\begin{eqnarray}
	\id{current}(\id{prec})&=&\overline{\id{prec}}^+;\id{prec}^+\cap I\\
	\id{initial}(\id{prec})&=&\id{prec}^+;\overline{\id{prec}}^+\cap I
\end{eqnarray}

@@

	Let $\mathbb T$ be the set of all conceivable traces.
	With \id{prec} being a history, we define a trace $(e_1,e_2)_\id{prec}$ as follows:
\begin{definition}[Traces]
\label{def:Traces}
\item $(e_1,e_2)_\id{prec}\ =\ e_1;\id{prec}^*\cap\id{prec}^*;e_2$
\end{definition}
	If it is obvious from the context which history ($\id{prec}$) is meant, we may write $(e_1,e_2)$ rather than $(e_1,e_2)_\id{prec}$.
	A trace algebra is a triple $\la\mathbb T,+,0\ra$,
	where the operator $+$ (concatenation) and empty trace $0$ satisfy the axioms of trace algebras:
\begin{eqnarray}
	x+(y+z)&=&(x+y)+z\label{axi:associative}\\
	0+x&=&x+0\label{axi:lr-identity}\\
	x+y\ =\ x+z&\Rightarrow&y=z\label{axi:left cancellation}\\
	x+z\ =\ y+z&\Rightarrow&x=y\label{axi:right cancellation}\\
	x+z\ =\ 0&\Rightarrow&x=0\label{axi:no inverse}
\end{eqnarray}
	In terms of universal algebra,
	equations~\ref{axi:associative} and~\ref{axi:lr-identity} state that a trace algebra is a {\em monoid}.
	Note that $+$ is {\em not} commutative.
	Equations~\ref{axi:left cancellation} and~\ref{axi:right cancellation} state that $+$ is {\em cancellative}.
	Equation~\ref{axi:no inverse} implies that there is no inverse.
	It makes the cancellative monoid into a trace algebra.

	For showing that definition~\ref{def:Traces} satisfies the axioms of trace algebra,
	we define the empty trace ($0$) and the trace concatenation ($+$) as $I_\mathbb E$ and $;$ (relation composition).
\begin{definition}[trace operators]
\item $0\ =\ I$
\item $x+y\ =\ x;y$
\end{definition}

@@

\section{State}
\label{sct:State}
	Typically, distributed information systems are stateful.
	Every event in the history of a DIS may cause a different (local) state.
	(The distributed nature of a DIS means that states are local.)
	Let $\mathbb S$ be the set of all conceivable states.
	So we need a function $\id{state}:\mathbb E\rightarrow\mathbb S$ that assigns one state to every event $e$ in history $\id{prec}$.
	State $\id{state}(e)$ represents the state immediately after $e$ has happened.
	If the history $\id{prec}$ is obvious from the context, we may write $\id{state}(e)$ rather than $\id{state}(e)$.
	We assume an empty state, $\emptyset$, which is the state immediately before an initial event has happened.
	The global state of the system, if relevant, can be computed by $\{\id{state}(e) | e\in\id{current}(\id{prec})\}$.
	Locally we work with one specific current event, $e$, with $\id{state}(e)$ representing the local current state.

@@

	We need operators to compute a new state ($s_2$) from a given state ($s_1$).
	In fact, we need two operators, $\oplus:\mathbb S\times\mathbb D\rightarrow\mathbb S$ and $\ominus:\mathbb S\times\mathbb S\rightarrow\mathbb D$, such that
\begin{equation}
	s_2\ =\ s_1\oplus(s_2\ominus s_1)
\label{req:oplus and ominus}
\end{equation}
	Let us call $s_2\ominus s_1$ the difference between states $s_2$ and $s_1$.
	The following equations represent the laws about state.
	They must be kept true all the time in a DIS.
\begin{eqnarray}
	\delta_\id{prec}(e_1,e_2)&=&\id{state}(e_2)\ominus\id{state}(e_1)\\
	e_1\ \id{prec}\ e_2&\Rightarrow&\id{state}(e_2)\ =\ \id{state}(e_1)\oplus\delta_\id{prec}(e_1,e_2)
\end{eqnarray}

@@

	A transaction is denoted as $\id{act}_{e_1,e_2,\Delta}$, which is a mapping from one DIS to another
\begin{equation}
\label{def:transaction}
\begin{array}{rl}
	\multicolumn{2}{l}{e_2\not\in\id{events}(\id{prec})\ \wedge\ (e_1,e_2)\in\id{prec}'^*\ \Rightarrow}\\
	&\id{act}_{e_1,e_2,\Delta}(\id{prec},\delta_\id{prec})\ =\ (\id{prec}',\delta'_{\id{prec}'})\\
	&\text{where}\\
	&\ \ \begin{array}[t]{rcll}
		\delta'_{\id{prec}'}(e,e')&=&\Delta&\text{if }(e,e')=(e_1,e_2)\\
		\delta'_{\id{prec}'}(e,e')&=&\delta_\id{prec}(e,e')&\text{otherwise}\\
		\end{array}
\end{array}
\end{equation}
	The state function can be computed recursively as follows:
\begin{equation}
\begin{array}{rcll}
	\id{state}(e)&=&\id{state}(e_1)\oplus\delta(e_1,e)&\text{if }(e_1,e)\in\id{prec}\\
	                       &=&\emptyset&\text{otherwise}
\end{array}
\end{equation}

	Now let us define the classical characteristic properties of transactions.
	Let us start with atomic.

	In his paper "Splitting Atoms Safely"~\cite{JONES2007109} Cliff Jones\footnote{%
	Jones is author of the Atomicity Manifesto and author of the formal method VDM.}
	describes the intended meaning of atomicity in natural language:
	``If an action is said to be executed “atomically”, it is assumed that it will not be affected by interference
	and that the environment will not be able to observe intermediate steps of the action in question.''
	Jones argues that interference must be tolerated, however.
	With shared state programs, he says, an action must achieve some required result even though its state can be changed by other interfering processes.

	The purpose of this paper is to propose a useful formal definition of atomicity.
	Atomicity expresses that a trace of multiple events can be treated as a single pair $\in\id{prec}$.
	To define this, we need an auxiliary function
	$\id{atomize}:\mathbb E\times\mathbb E\times\mathbb H\rightarrow\mathbb H$,
	which replaces a trace $(e_1,e_2)_\id{prec}$ by the pair $(e_1,e_2)$
	in a history $\id{prec}$.
\begin{eqnarray}
	\begin{array}[t]{r@{}l@{}l}
	\multicolumn{3}{l}{\id{atomize}(e_1,e_2,\id{prec})}\\
	   \ =\ \id{prec}-&(&\{ (e',e)\in\id{prec}| e_1\ \id{prec}^*\ e\ \wedge\ e\ \id{prec}^*\ e_2\}\ \cup\\
			  & &\{ (e,e')\in\id{prec}| e_1\ \id{prec}^*\ e\ \wedge\ e\ \id{prec}^*\ e_2\}\\
			  &)\\
		\cup&&\{ (e_1,e_2) \}\\
	\end{array}
\end{eqnarray}

@@

	Now let us turn to consistency.
	This property means that a transaction preserves truth of the state.
	We need a predicate $\mathcal P\{\mathbb S\}$ to express that a state is consistent.
\begin{definition}[Consistency]
\item A transaction $\id{act}_{e_1,e_2,\Delta}$ is consistent w.r.t. $\mathcal P$ iff:
\[\mathcal P\{\id{state}(e_1)\}\ \Rightarrow\ \mathcal P\{\id{state}(e_2)\}\]
\label{def:Consistency}
\end{definition}

	The third property is isolation.
	A transaction $\id{act}(e_1,e_2)$ is isolated, i.e. performed in isolation, if it does not affect any other transaction in the system.
	This means that the only event in trace $(e_1,e_2)_\id{prec}$ that precedes events outside that trace is $e_1$.
	Likewise, the only event in trace $(e_1,e_2)_\id{prec}$ that is preceded by events outside that trace is $e_2$.
	We can now define isolation as the property that the traces of the ``atomized'' history
	are the same as the atomization of all traces in that history.
\begin{definition}[Isolation]
\item A transaction $\id{act}_{e_1,e_2,\Delta}$ works isolated in history $\id{prec}$ iff:
\label{def:Isolation}
\begin{eqnarray}
	(\id{atomize}(e_1,e_2,\id{prec}))^*&=&\id{atomize}(e_1,e_2,\id{prec}^*)
\end{eqnarray}
\end{definition}
